Estimating demographic bias on tests of children's early vocabulary

### 2022-02-01 paper/cogsci.Rmd
minor edits, fig tweaks

One challenge in comprehensive assessment of these differences is that is difficult to get a complete measure of young children's language skill. Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

@@

The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children. 
Due to sparse data for many categories, we binarized participants' race andethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there is important variation between groups that this will fail to capture. 
Data for the maternal education analysis included CDI data from children whose mothers had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).

@@

First, we examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
Then we fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
Next, we re-examine the size of demographic effects after pruning biased items from the CDI:WS using different criteria, and recommend a threshold.
Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs (Figures 3 and 4) -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This pattern makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model. Words at the top are more well-known by females, while those at the bottom are more known for males."}
knitr::include_graphics("figs/smGLIMMER_combo.pdf")
```

@@

pr1 <- pdat %>% filter(Dimension=="sex") %>%
  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
  geom_point() + #theme(legend.position = "bottom") + 
  labs(color="Excluded items") + ylim(-.7, -.4) + ggtitle("Sex") +
  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")

pr2 <- pdat %>% filter(Dimension=="maternal ed") %>%
  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
  geom_point() + 
  labs(color="Excluded items") + ylim(-.3, -.1) + ggtitle("Maternal Education") +
  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")

pr3 <- pdat %>% filter(Dimension=="race") %>%
  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
  geom_point() + 
  labs(color="Excluded items") + ylim(.35, .55) + ggtitle("Race") +
  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")

@@

#white=.48
#highSES=0.20
#female=0.57

@@

```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic."}
knitr::include_graphics("figs/extrema2p25sd.pdf")
```

>>>>

One challenge in comprehensive assessment of these differences is that it is difficult to get a complete measure of young children's language skill. Long recordings are prohibitively difficult to collect and transcribe, and yet any short recording (e.g., a 1-hour play session) will elicit only a small proportion of the words and constructions that children know. 
Thus, researchers of early word learning have constructed tests with hundreds of words, intentionally oversampling words that are more likely to be known by young children.

@@

The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
The analysis of race-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children. 
Due to sparse data for many categories, we binarized participants' race and ethnicity as White (`r eth2_tab["White"]`) or Non-white (`r eth2_tab["Nonwhite"]`), recognizing that there is important variation between groups that this will fail to capture. 
Data for the maternal education analysis included CDI data from children whose mothers had the following levels of education: `r momed_tab["Primary"]` with no more than primary school education, `r momed_tab["Some Secondary"]` with some secondary school, `r momed_tab["Secondary"]` with no more than secondary school, `r momed_tab["Some College"]` with some college, `r momed_tab["College"]` with no more than a college degree, `r momed_tab["Some Graduate"]` with no more than some graduate school, and `r momed_tab["Graduate"]` with a graduate degree.
Again due to data sparsity, we binarized the `r ses_tab["high"]` children whose mothers had at least some college or more as high maternal education (high-ME), and those whose mothers had at most high school (`r ses_tab["low"]` children) as low maternal education (low-ME).

@@

First, we examine the size of demographic effects on language ability in a baseline Rasch model fitted without regard to demographic group.
Then we fit a multigroup Rasch model for each demographic factor, and characterize the between-group differences in item difficulties.
Next, we re-examine the size of demographic effects after pruning biased items from the CDI:WS using different criteria, and recommend a threshold.
Finally, we measure the strength of association between sex-related differences in language input and the degree of sex-related DIF.
The scripts and models needed to reproduce our analyses are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a).

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of the sex GLIMMER (Fig. 2, left): at the top, "vagina"^[Parents using the CDI are instructed to check genital items if the child produces whatever word the family uses.], "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs -- and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This pattern makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plots of a sample of CDI:WS words from the sex bias model (left), the maternal education model (center), and the race model (right). Words at the top are more well-known by one group (e.g., females, for sex), while those at the bottom are more known by the other group (e.g., males), and the selection in the middle show roughly the median difficulty difference for those groups."}
knitr::include_graphics("figs/smGLIMMER_combo.pdf")
```

@@

pr1 <- pdat %>% filter(Dimension=="sex") %>%
  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
  geom_point() + theme_bw(base_size=9) +
  labs(color="Excluded items") + ylim(-.7, -.4) + ggtitle("Sex") +
  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")

pr2 <- pdat %>% filter(Dimension=="maternal ed") %>%
  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
  geom_point() + theme_bw(base_size=9) +
  labs(color="Excluded items") + ylim(-.3, -.1) + ggtitle("Maternal Education") +
  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")

pr3 <- pdat %>% filter(Dimension=="race") %>%
  ggplot(aes(x=SDthreshold, y=beta, color=extreme)) + 
  geom_point() + theme_bw(base_size=9) +
  labs(color="Excluded items") + ylim(.35, .55) + ggtitle("Race") +
  xlab("Exclusion Threshold (SDs)") + ylab("Coefficient (beta)")

@@

#white=.48
#highME=0.20
#female=0.57

@@

```{r biased-words, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=3.2, set.cap.width=T, num.cols.cap=1, fig.cap = "The 59 items showing extreme bias (difficulty difference $>2.25$ SD) for one or more demographic. The 19 items in bold are those showing a difference $>3$ SD."}
knitr::include_graphics("figs/extrema2p25sd_3sd_bold.pdf")
```
